---
title: "STEP_4_5_AIM_1_JIVE_validation.Rmd"
output: html_document
date: "2023-10-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(robencla)
library(survival)
library(survminer)

```

## JIVE Validation

```{r}


# Take the data set, and filter out pair genes that are not present.

clean_genepairs_list <- function(genepairs, dat_x) {
  genepairs_clean <- list()
  i <- 1
  for (pi in names(genepairs)) {
    genes <- genepairs[[pi]]
    missing_genes <- unique( genes [! genes %in% colnames(dat_x)] )
    print(paste0(pi, '  ', missing_genes))
    newpairlist <- c()
    for (j in seq.int(from=1,to=length(genes), by=2)) {
      if ( (genes[j] %in% missing_genes) | (genes[j+1] %in% missing_genes) ) {
        # then don't add them!
        print(paste0("removing from pair list:  ", genes[j], ' ', genes[j+1]))
      } else {
        newpairlist <- c(newpairlist, genes[j], genes[j+1])
      }
    }
    genepairs_clean[[pi]] <- newpairlist
  }
  return(genepairs_clean)
}

# pairlist is a list of vectors (paired terms)
# n is the number of pairs for each entry desired

shorter_plist <- function(pairlist, n) {
  newlist <- list()
  for (ni in names(pairlist)) {
    newlist[[ni]] <- pairlist[[ni]][1:(2*n)]
  }
  return(newlist)  
}



```


First training the full model with all data.
```{r}

library(readr)
library(robencla)
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)

dat_f <- read_csv('../data/jive_training_array_data_F.csv.gz')
dat_f <- dat_f[,-1] # remove the first column

val_f1 <- read_csv('../data/Array Data/jive_validation_16011_F.csv.gz')
val_f1$Survival <- as.numeric(val_f1$Survival) * 30.437

val_f2 <- read_csv('../data/Array Data/validation_array_gse108474_F.csv.gz')
val_f2$Survival <- as.numeric(val_f2$Survival) * 30.437

# survival was in days
val_f3 <- read_csv('../data/Array Data/jive_validation_13041_ex3_F_v2.csv.gz')
val_f3$Censored <- val_f3$Censored-1

load("../results/results_no_array_min/females_genepairs.rda")
lapply(genepairs, length)

# "fix" a couple columns
gidx <- which(colnames(val_f1) == 'LGALS2')
colnames(val_f1)[gidx] <- 'LGALS3'
gidx <- which(colnames(val_f1) == 'GFOD3P')
colnames(val_f1)[gidx] <- 'TP73-AS1'
gidx <- which(colnames(val_f2) == 'SubjectID')
colnames(val_f2)[gidx] <- 'SampleID'  # like the other data sets
```


```{r}

gpairs <- shorter_plist(genepairs, 10)

lapply(gpairs, function(x) x[! (x %in% colnames(dat_f ))] )
lapply(gpairs, function(x) x[! (x %in% colnames(val_f1))] ) # missing "FABP5"  "CD24"   "LGALS3"
lapply(gpairs, function(x) x[! (x %in% colnames(val_f2))] )
lapply(gpairs, function(x) x[! (x %in% colnames(val_f3))] )

# clean the pairs
genepairs_cl <- clean_genepairs_list(gpairs, val_f1)  # only one with missing genes

gs <- c(unlist(unique(genepairs_cl)), "SampleID", "Sex", "Survival", "Censored")

# combining the validation data
val_dat <- rbind(val_f1[,gs], val_f2[,gs])
val_dat <- rbind(val_dat, val_f3[,gs])
#val_dat <- val_f2[,gs]

# what if ...
dat_f$ClusterLabel2 <- ifelse(dat_f$ClusterLabel == 'cluster3', yes = 'C3', no='notC3')
c3_pairs <- list()
c3_pairs[['C3']] <- genepairs_cl$cluster3
c3_pairs[['notC3']] <- c(genepairs_cl$cluster1, genepairs_cl$cluster2, genepairs_cl$cluster4, genepairs_cl$cluster5)


# xgboost parameters to pass to each sub-classifier in the ensembles
# xgboost parameters to pass to each sub-classifier in the ensembles
params1 <- list(max_depth=12,   # "height" of the tree, 6 is actually default. I think about 12 seems better.  (xgboost parameter)
               eta=0.45,        # this is the learning rate. smaller values slow it down, more  conservative   (xgboost parameter)
               nrounds=24,     # number of rounds of training, lower numbers less overfitting (potentially)  (xgboost parameter)
               early_stopping_rounds=2,
               nthreads=8,     # parallel threads
               gamma=0.2,      # Minimum loss req'd to again partition a leaf node. higher number ~ more conservative (xgboost)
               lambda=2.5,     # L2 regularization term on weights, higher number ~ more conservative (xgboost parameter)
               alpha=0.25,      # L1 regularization term on weights. higher number ~ more conservative (xgboost parameter)
               verbose=0,
               train_perc=0.8,
               combine_function='median',
               size=11
)


mod_f <- robencla::Robencla$new('f_model')


mod_f$train(data_frame=dat_f,
            label_name='ClusterLabel', #'ClusterLabel2',
            sample_id = 'Barcode',
            drop_list = c('Sex','ClusterLabel2'),
            data_mode=c('pairs'),    # pairs, allpairs, sigpairs, quartiles, tertiles, binarize, ranks, original #
            signatures=NULL,         # 
            pair_list=genepairs_cl,  # c3_pairs   # subset to these genes.
            params=params1
            )


# run the model
mod_f$predict(data_frame = val_dat, 
              sample_id = 'SampleID',
              drop_list = c('Sex','Survival', 'Censored'))



table(mod_f$results()$BestCall)
# rembrandt
#      37       37        7       60 
      
# get them reordered
### selected females in order of results
rownames(val_dat) <- val_dat$SampleID
pheno3_reorg <- val_dat[mod_f$results()$SampleIDs, c("SampleID","Sex","Survival","Censored")]

# now bind in the results to the phenotype data
resdf <- cbind(pheno3_reorg, mod_f$results())
#resdf <- na.omit(resdf)
resdf$Survival <- as.numeric(resdf$Survival)

resdf$C3 <- unlist(resdf$C3)
resdf$notC3 <- unlist(resdf$notC3)
resdf$cluster1 <- unlist(resdf$cluster1)
resdf$cluster2 <- unlist(resdf$cluster2)
resdf$cluster3 <- unlist(resdf$cluster3)
resdf$cluster4 <- unlist(resdf$cluster4)
resdf$cluster5 <- unlist(resdf$cluster5)

write_csv(resdf, file='../results/validation_F_all_results_genepairs_size12.csv')

modfit <- survfit(Surv(Survival,Censored)~BestCalls,data=resdf)
ggsurvplot(modfit, pval = T,xlim=c(0,1700))
ggsave('../figures/validation_survival_curve_F_genepairs_size12.pdf')

imp_list <- mod_f$importance()
save(imp_list, file='../results/female_validation_pairs_importance_step4_5_size12.rda')
df1 <- data.frame(Top10=1:10, Cluster='cluster1',imp_list[['cluster1']][1:10,])
df2 <- data.frame(Top10=1:10, Cluster='cluster2',imp_list[['cluster2']][1:10,])
df3 <- data.frame(Top10=1:10, Cluster='cluster3',imp_list[['cluster3']][1:10,])
df4 <- data.frame(Top10=1:10, Cluster='cluster4',imp_list[['cluster4']][1:10,])
df5 <- data.frame(Top10=1:10, Cluster='cluster5',imp_list[['cluster5']][1:10,])
impdf <- rbind(df1[,c(1,2,4)],df2[,c(1,2,4)],df3[,c(1,2,4)],df4[,c(1,2,4)],df5[,c(1,2,4)])
colnames(impdf) <- c('Top10','Cluster','Med_Info_Gain')
allimp <- rbind(df1,df2,df3,df4,df5)
write.csv(allimp, file='../results/female_validation_importance_table.csv')

save(mod_f, file='../models/females_tcga_array_step4_5_size12.rda')

tcga_train_dat <- mod_f$train_data
tcga_train_labels <- mod_f$train_label
val_test_dat <- mod_f$test_data

val_test_dat$SampleID <- resdf$SampleID
val_test_dat$Sex <- resdf$Sex
val_test_dat$Survival <- resdf$Survival
val_test_dat$ClusterLabel <- resdf$BestCalls

save(resdf, val_test_dat, tcga_train_dat, tcga_train_labels, genepairs_cl, file='../results/female_validation_test_data.rda')

# scores for the samples called cluster3
boxplot(resdf$cluster3~as.factor(resdf$BestCalls))



#boxplot(resdf$Survival ~ resdf$BestCall, 
#        xlab='Model Prediction', ylab='Survival (days)')

# extra check
#c3_call <- ifelse(resdf$BestCall == 'cluster3', yes=1, no=0) #ifelse(mod_f$pred_table$cluster3 > 0.8, yes=1, no=0)

#boxplot(resdf$Survival ~ c3_call, 
#        xlab='Model Prediction', ylab='Survival (days)')

```





test the validation array annotations for accuracy!

```{r}

# test the validation array annotations for accuracy!


load('../results/results_no_array_min/female_validation_test_data.rda')

# xgboost parameters to pass to each sub-classifier in the ensembles
# xgboost parameters to pass to each sub-classifier in the ensembles
params1 <- list(max_depth=12,   # "height" of the tree, 6 is actually default. I think about 12 seems better.  (xgboost parameter)
               eta=0.45,        # this is the learning rate. smaller values slow it down, more  conservative   (xgboost parameter)
               nrounds=24,     # number of rounds of training, lower numbers less overfitting (potentially)  (xgboost parameter)
               early_stopping_rounds=2,
               nthreads=8,     # parallel threads
               gamma=0.2,      # Minimum loss req'd to again partition a leaf node. higher number ~ more conservative (xgboost)
               lambda=2.5,     # L2 regularization term on weights, higher number ~ more conservative (xgboost parameter)
               alpha=0.25,      # L1 regularization term on weights. higher number ~ more conservative (xgboost parameter)
               verbose=0,
               train_perc=0.8,
               combine_function='median',
               size=11
)


mod_f <- robencla::Robencla$new('f_model')


mod_f$train(data_frame=val_test_dat,
            label_name='ClusterLabel', #'ClusterLabel2',
            sample_id = 'SampleID',
            drop_list = c('Sex','Survival','Censored'),
            data_mode=c('pairs'),    # pairs, allpairs, sigpairs, quartiles, tertiles, binarize, ranks, original #
            signatures=NULL,         # 
            pair_list=genepairs_cl,  # c3_pairs   # subset to these genes.
            params=params1
            )


# run the model
mod_f$predict(data_frame = tcga_train_dat, 
              sample_id  = 'Barcode',
              label_name = 'ClusterLabel',
              drop_list  = 'Sex')



table(mod_f$results()$BestCall, mod_f$test_label)
## check results

results <- mod_f$results()
results$TestLabel <- mod_f$test_label

write_csv(results, file='../results/females_val_dat_trained_pred_on_TCGA_results_table_step4_5.csv')

metrics <- mod_f$classification_metrics(these_labels = mod_f$test_label, these_calls = mod_f$results()$BestCall,use_cv_results = F)
write_csv(metrics, file='../results/females_val_dat_trained_pred_on_TCGA_metrics_step4_5.csv')

```

size 12, validation training, predicting back to TCGA

           cluster1 cluster2 cluster3 cluster4 cluster5
  cluster1       41        4        0        2        0
  cluster2        0       19        0        0        0
  cluster3        0        0       11        0        0
  cluster4        0        1        0       18        0
  cluster5        0        0        3        0       40









First training the full model with all male data.
```{r}

library(readr)
library(robencla)
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2)

#dat_m <- read_csv('../data/jive_training_array_data_M.csv.gz')
#dat_m <- dat_m[,-1] # remove the first column
# val_m1 <- read_csv('../data/Array Data/jive_validation_16011_M.csv.gz')
# val_m1$Survival <- as.numeric(val_m1$Survival) * 30.437
# val_m2 <- read_csv('../data/Array Data/validation_array_gse108474_M.csv.gz')
# val_m2$Survival <- as.numeric(val_m2$Survival) * 30.437
# # survival was in days
# val_m3 <- read_csv('../data/Array Data/jive_validation_13041_ex3_M_v2.csv.gz')
# val_m3$Censored <- val_m3$Censored-1
#load("../results/males_genepairs.rda")
#lapply(genepairs, length)
# "fix" a couple columns
#gidx <- which(colnames(val_m1) == 'LGALS2')
#colnames(val_m1)[gidx] <- 'LGALS3'
#gidx <- which(colnames(val_m1) == 'GFOD3P')
#colnames(val_m1)[gidx] <- 'TP73-AS1'
#gidx <- which(colnames(val_m2) == 'SubjectID')
#colnames(val_m2)[gidx] <- 'SampleID'  # like the other data sets

# see data set up
load('../data/M_processed_data.rda')
load('../data/m_genepairs.rda')

# removing signature genes that are not available in our data sets
male.jive.gene.list <- readRDS("../data/Males_jive_cluster_genes.rds") %>% dplyr::filter(value %in% colnames(val_m))
sigs_m <- list("C1"=male.jive.gene.list$value[male.jive.gene.list$cluster.group == 'cluster1'],
               "C2"=male.jive.gene.list$value[male.jive.gene.list$cluster.group == 'cluster2'],
               "C3"=male.jive.gene.list$value[male.jive.gene.list$cluster.group == 'cluster3'],
               "C4"=male.jive.gene.list$value[male.jive.gene.list$cluster.group == 'cluster4'],
               "C5"=male.jive.gene.list$value[male.jive.gene.list$cluster.group == 'cluster5']
)


```


```{r}

gpairs <- shorter_plist(m_genepairs$val, 24)

lapply(gpairs, function(x) x[! (x %in% colnames(dat_m))] )
lapply(gpairs, function(x) x[! (x %in% colnames(val_m))] ) # missing "ZNF710-AS1" in C4 and "NEFL" in C5

# clean the pairs
genepairs_cl <- clean_genepairs_list(gpairs, val_m)  # only one with missing genes

gs <- c(unlist(unique(genepairs_cl)), "SampleID", "Sex", "Survival", "Censored")

# xgboost parameters to pass to each sub-classifier in the ensembles
# xgboost parameters to pass to each sub-classifier in the ensembles
params1 <- list(max_depth=12,   # "height" of the tree, 6 is actually default. I think about 12 seems better.  (xgboost parameter)
               eta=0.45,        # this is the learning rate. smaller values slow it down, more  conservative   (xgboost parameter)
               nrounds=24,     # number of rounds of training, lower numbers less overfitting (potentially)  (xgboost parameter)
               early_stopping_rounds=2,
               nthreads=8,     # parallel threads
               gamma=0.2,      # Minimum loss req'd to again partition a leaf node. higher number ~ more conservative (xgboost)
               lambda=2.5,     # L2 regularization term on weights, higher number ~ more conservative (xgboost parameter)
               alpha=0.25,      # L1 regularization term on weights. higher number ~ more conservative (xgboost parameter)
               verbose=0,
               train_perc=0.8,
               combine_function='median',
               size=11
)

mod_m <- robencla::Robencla$new('m_model')

mod_m$train(data_frame=dat_m,
            label_name='ClusterLabel', #'ClusterLabel',
            sample_id = 'Barcode',
            drop_list = c('Sex'),
            data_mode=c('pairs'),    # pairs, allpairs, sigpairs, quartiles, tertiles, binarize, ranks, original #
            signatures=NULL,         # 
            pair_list=genepairs_cl, #genepairs_cl, # c3_pairs  # subset to these genes.
            params=params1
            )


# run the model
mod_m$predict(data_frame = val_m,
              sample_id = 'Barcode',
              drop_list = c('Sex','Survival', 'Censored'))


table(mod_m$results()$BestCall)

#with 24 pairs from val
#cluster1 cluster2 cluster3 cluster4 cluster5 
#      41      150       56       39      250 

# 12 from rna
#cluster1 cluster2 cluster3 cluster4 cluster5 
#      32      165       85        4      250 

# get the results table
resdf <- mod_m$results()
# add in the survival and censored
resdf$Survival <- mod_m$test_data$Survival
resdf$Censored <- mod_m$test_data$Censored

# convert scores to numerics
resdf$cluster1 <- unlist(resdf$cluster1)
resdf$cluster2 <- unlist(resdf$cluster2)
resdf$cluster3 <- unlist(resdf$cluster3)
resdf$cluster4 <- unlist(resdf$cluster4)
resdf$cluster5 <- unlist(resdf$cluster5)

modfit <- survfit(Surv(Survival,Censored)~BestCalls,data=resdf)
ggsurvplot(modfit, pval = T,xlim=c(0,1700))


rclin <- read.csv('../data/Array Data/validation_array_resources/GSE108475_REMBRANDT_clinical.data.csv')

df <- inner_join(resdf, rclin, join_by('SampleIDs'=='SUBJECT_ID'))

table(df$BestCalls, df$DISEASE_TYPE)




# get them reordered
### selected females in order of results
#rownames(val_dat) <- val_dat$SampleID
#pheno3_reorg <- val_dat[mod_m$results()$SampleIDs, c("SampleID","Sex","Survival","Censored")]

# now bind in the results to the phenotype data
#resdf <- cbind(pheno3_reorg, mod_m$results())
#resdf <- na.omit(resdf)
#resdf$Survival <- as.numeric(resdf$Survival)

#resdf$C3       <- unlist(resdf$C3)
#resdf$notC3    <- unlist(resdf$notC3)
#resdf$cluster1 <- unlist(resdf$cluster1)
#resdf$cluster2 <- unlist(resdf$cluster2)
#resdf$cluster3 <- unlist(resdf$cluster3)
#resdf$cluster4 <- unlist(resdf$cluster4)
#resdf$cluster5 <- unlist(resdf$cluster5)
# 
# modfit <- survfit(Surv(Survival,Censored)~BestCalls,data=resdf)
# ggsurvplot(modfit, pval = T,xlim=c(0,1700))
# ggsave('../figures/males_validation_survival_curve_all_size24.pdf')
# 
# ggsave('../figures/males_validation_survival_curve_all_size24_binary_C5.pdf')
# 
# # scores for the samples called cluster3
# boxplot(resdf$cluster5~as.factor(resdf$BestCalls))
# 
# # get the importance
# imp_list <- mod_m$importance()
# save(imp_list, file='../results/male_validation_pairs_importance_step4_5_size24.rda')
# 
# df1 <- data.frame(Top10=1:10, Cluster='cluster1',imp_list[['cluster1']][1:10,])
# df2 <- data.frame(Top10=1:10, Cluster='cluster2',imp_list[['cluster2']][1:10,])
# df3 <- data.frame(Top10=1:10, Cluster='cluster3',imp_list[['cluster3']][1:10,])
# df4 <- data.frame(Top10=1:10, Cluster='cluster4',imp_list[['cluster4']][1:10,])
# df5 <- data.frame(Top10=1:10, Cluster='cluster5',imp_list[['cluster5']][1:10,])
# impdf <- rbind(df1[,c(1,2,4)],df2[,c(1,2,4)],df3[,c(1,2,4)],df4[,c(1,2,4)],df5[,c(1,2,4)])
# colnames(impdf) <- c('Top10','Cluster','Med_Info_Gain')
# allimp <- rbind(df1,df2,df3,df4,df5)
# write.csv(allimp, file='../results/male_validation_importance_table.csv')
# 
# save(mod_m, file='../models/males_tcga_array_step4_5_size24.rda')
# 
# write_csv(resdf, file='../results/males_validation_all_results_step4_5_size24.csv')
# 
# 
# 
tcga_train_dat    <- mod_m$train_data
tcga_train_dat$ClusterLabels <- mod_m$train_label
val_test_dat      <- mod_m$test_data
 
val_test_dat$Barcode <- mod_m$test_data$Barcode
val_test_dat$Survival <- mod_m$test_data$Survival
val_test_dat$ClusterLabel <- mod_m$results()$BestCalls

#save(val_test_dat, tcga_train_dat file='../data/val_test_dat.rda')
# 
# save(resdf, val_test_dat, tcga_train_dat, tcga_train_labels, genepairs_cl, c3_pairs, file='../results/male_validation_test_data_binary_C5.rda')
# 
# # scores for the samples called cluster3
# boxplot(resdf$Survival~as.factor(resdf$BestCalls))
# 
# 
# #boxplot(resdf$Survival ~ resdf$BestCall, 
#        xlab='Model Prediction', ylab='Survival (days)')

# extra check
#c3_call <- ifelse(resdf$BestCall == 'cluster3', yes=1, no=0) #ifelse(mod_f$pred_table$cluster3 > 0.8, yes=1, no=0)

#boxplot(resdf$Survival ~ c3_call, 
#        xlab='Model Prediction', ylab='Survival (days)')

```




test the validation array annotations for accuracy!

```{r}

# test the validation array annotations for accuracy!


load('../data/val_test_dat.rda')

gpairs <- shorter_plist(m_genepairs$val, 24)

lapply(gpairs, function(x) x[! (x %in% colnames(val_test_dat))] ) # missing "ZNF710-AS1" in C4 and "NEFL" in C5

# clean the pairs
genepairs_cl <- clean_genepairs_list(gpairs, val_test_dat)  # only one with missing genes

# xgboost parameters to pass to each sub-classifier in the ensembles
# xgboost parameters to pass to each sub-classifier in the ensembles
params1 <- list(max_depth=12,   # "height" of the tree, 6 is actually default. I think about 12 seems better.  (xgboost parameter)
               eta=0.3,        # this is the learning rate. smaller values slow it down, more  conservative   (xgboost parameter)
               nrounds=24,     # number of rounds of training, lower numbers less overfitting (potentially)  (xgboost parameter)
               early_stopping_rounds=2,
               nthreads=8,     # parallel threads
               gamma=1.2,      # Minimum loss req'd to again partition a leaf node. higher number ~ more conservative (xgboost)
               lambda=2.5,     # L2 regularization term on weights, higher number ~ more conservative (xgboost parameter)
               alpha=1.25,      # L1 regularization term on weights. higher number ~ more conservative (xgboost parameter)
               verbose=0,
               train_perc=0.8,
               combine_function='median',
               size=11
)


mod_m <- robencla::Robencla$new('m_model')


mod_m$train(data_frame=val_test_dat,
            label_name='ClusterLabel', #'ClusterLabel2',
            sample_id = 'SampleID',
            drop_list = c('Sex','Survival','Censored'),
            data_mode=c('pairs'),    # pairs, allpairs, sigpairs, quartiles, tertiles, binarize, ranks, original #
            signatures=NULL,         # 
            pair_list=genepairs_cl,  # c3_pairs   # subset to these genes.
            params=params1
            )


# run the model
mod_m$predict(data_frame = tcga_train_dat, 
              sample_id  = 'Barcode',
              label_name = 'ClusterLabel')



table(Pred=mod_m$results()$BestCall, True=mod_m$test_label)
## check results


val_dat <- mod_m$train_data
save(val_dat, file='../data/validation_data_with_pred_labels.rda')

results <- mod_m$results()
results$TestLabel <- mod_m$test_label

write_csv(results, file='../results/males_val_dat_trained_pred_on_TCGA_results_table_step4_5.csv')

metrics <- mod_m$classification_metrics(these_labels = mod_m$test_label, these_calls = mod_m$results()$BestCall, use_cv_results = F)
write_csv(metrics, file='../results/males_val_dat_trained_pred_on_TCGA_metrics_step4_5.csv')



```

On the set 1 - 24 pairs.

           cluster1 cluster2 cluster3 cluster4 cluster5
  cluster1       29        0        0        1        0
  cluster2        2       80        1        1        2
  cluster3        1        0       24        1        0
  cluster4        6        5        8       23        3
  cluster5        0        0        2        0       31


saved with the val_dat_with_pred_labels
          True
Pred       cluster1 cluster2 cluster3 cluster4 cluster5
  cluster1       26        2        0        2        1
  cluster2        3       77        1        2        3
  cluster3        1        0       25        1        0
  cluster4        8        6        3       21        0
  cluster5        0        0        6        0       32










